---
title: "Next Word Prediction"
author: "Nikhil Parimi"
date: "2024-12-27"
output: ioslides_presentation
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Goals and Aims

This presentation will give a brief run down of the work done throughout the project and what went into building a word prediction algorithm as well as its deployment.

The project can be viewed at:

- [https://nikhilparimi.shinyapps.io/word_prediction/](https://nikhilparimi.shinyapps.io/word_prediction/)

The code for the project is located at:

- [https://github.com/nikhilparimi/DS-Capstone](https://github.com/nikhilparimi/DS-Capstone)

In the next few slides we will go over:

- Data Processing and Cleaning
- Predictive Algorithm
- Deployment of application

## Data Processing and Cleaning

For this model I have sampled 15% of the data (from the blog, news and twitter datasets), further cleaning and tokenising them, resulting in a sample dataset of 500,000 lines and 10.2 million words!

This cleaned dataset was then used as input in the `unnest_ngrams()` function which returned all possible n-grams for a particular dataset for a specified value of n. For the purposes of my model I created n-grams in the range [1-4]. The generated n-grams were then restructured to the form `phrase` and `frequency` so that further processing could be done.

After this I did some further cleaning of the generated n-grams as the data generated was too large to deal with so I manually filtered the frequency cut off for each of the n-grams

## Predictive Algorithm

The string for which a prediction needs to be made is passed into a recursive backoff algorithm. **Essentially,** the algorithm checks the string that most closely matches the inputted string in the highest n-gram (n=4). If it is not found here, it recursively moves down n-grams until it reaches the base case.

```{r, eval = FALSE, echo = TRUE}
backoff_prediction <- function(x) {
  # ... input is tokenised here
  if (length(words) >= 3) {
    # ... processing / search is done here
    if (nrow(match_exact) > 0) {
      return(match_exact$word4[1]) # Returns word prediction
    }
    else {
      # ...
      backoff_prediction(tail_words) # Recursive function call
    }
  # ... Code continued
```


## Deployment Of Application and User Interface

The completed application was then deployed onto shinyapps.io using `rsconnect::deployApp('Word_Prediction')`. The application then predicts and displays a word based on the user's input.

Notes:

- This project was a deeply fulfilling journey where I got to learn a lot about data science and applying it in R.
- I want to thank you for reading my presentation on my application.
